---
title: "Reuse with R, R Markdown & git)"
author: "Paul Sharpe"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  beamer_presentation:
    fig_caption: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::read_chunk('external.R')
```

## R Markdown

* Quick (like a wiki) communication tool PDF, Word, HTML ...
* ... [Source that generated this presentation](https://github.com/earcanal/documents/Reuse.Rmd)

## Git(hub)

* De facto open collaborative development tool
* Feature rich, but you can get by with limited vocabulary: `clone`, `add`, `commit`, `push`, `diff`
* Commit messages as [project history metadata](https://github.com/earcanal/rsvp-task/commits/master)

## Don't Repeat Yourself (DRY)

* Minimise viral code!
* Reuse with functions, packages, and ...

## ... with knitr

The following number came from the chunk defined as `# ---- external-chunk ----` in [external.R](https://github.com/earcanal/documents/external.R): 

```{r external-chunk}
```

* Can't use symlinks due to [an issue in knitr](https://github.com/rstudio/rmarkdown/issues/1508)
* [include chunks with `knitr::read_chunk()`](https://yihui.name/knitr/demo/externalization/)
* Bug/feature: doesn't modify Global Environment because knitting is done in R sessions which are discarded
* Open and source (pre-processing) code in a different project to set correct working directory (yay, debugger!)
* Preregistration: open `.R`, embargo `.Rmd`